String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-12 2023-1-19/sections/preliminaries.tex
section 5 and 6

\medskip
\noindent \emph{Strings and languages.}
%\paragraph*{Strings and languages.}
A string over $\Sigma$ is a (possibly empty) sequence of elements from $\Sigma$,
denoted by $u, v, w, \ldots$. An empty string is denoted by $\varepsilon$. We use $\Sigma_\varepsilon$ to denote $\Sigma \cup \{\varepsilon\}$. We write $\Sigma^*$ (resp., $\Sigma^+$) for the set of all (resp. nonempty) strings over $\Sigma$.
%Let 
For a string $u$, we use $|u|$ to denote the number of letters in $u$. In particular, $|\varepsilon|=0$. 
Moreover, for $a \in \Sigma$, let $|u|_a$ denote the number of occurrences of $a$ in $u$. 
Assume $u=a_0\cdots a_{n-1}$ is nonempty and $i<j \in [0,n-1]$. %Then a \emph{position} of $u$ is a number $i \in [|u|]$ (Note that the first position is $1$, instead of  0). In addition, 
We let $u[i]$ denote $a_i$ and $u[i,j]$ for the substring %of $u$ starting from $i$ and ending with $j$ (i.e., 
$a_i\cdots a_j$. 
%\tl{check later for consistency}\zhilin{the indices start from 0, to be consistent with the semantics of substring and indexof}\zhilin{i do use it in section 4}
%
Let $u, v$ be two strings. We use $u \cdot v$ to denote the \emph{concatenation} of $u$ and $v$. A language $L$ over $\Sigma$ is a subset of strings.  
Let $L_1, L_2$ be two languages. Then the concatenation of $L_1$ and $L_2$, denoted by $L_1 \cdot L_2$,  is defined as $\{u \cdot v \mid u \in L_1, v \in L_2\}$. The union (resp. intersection) of $L_1$ and $L_2$, denoted by $L_1 \cup L_2$  (resp. $L_1 \cap L_2$), is defined as $\{u \mid u \in L_1 \mbox{ or } u \in L_2\}$ (resp. $\{u \mid u \in L_1 \mbox{ and } u \in L_2\}$). The complement of $L_1$, denoted by $\overline{L_1}$, is defined as $\{u \mid u \in \Sigma^*, u \not \in L_1\}$. 
The difference of $L_1$ and $L_2$, denoted by $L_1 \setminus L_2$, is defined as $L_1 \cap \overline{L_2}$. 
For a language $L$ and $n \in \Nat$, we define $L^n$ inductively as follows: $L^0= \{\varepsilon\}$, $L^{n+1} = L \cdot L^n$ for every $n \in \Nat$. 
Finally, define $L^* = \bigcup \limits_{n \in \Nat} L^n$.
%, that is, the string $w$ such that $|w|= |u| + |v|$ and for each $i \in [0, |u|-1]$, $w[i]= u[i]$, and for each $i \in [0,|v|-1]$, $w[|u|+i]=v[i]$.

@@

\medskip
\noindent 
\emph{Linear integer arithmetic and Parikh images.}
We will also use standard quantifier-free/existential \emph{linear integer arithmetic} (LIA) formulae, which are typically ranged over by $\phi, \varphi$, etc. 
For a set $\mathfrak{X}$ of variables, we use $\Phi(\mathfrak{X})$ to denote the set of quantifier-free LIA formulas whose free variables are from $\mathfrak{X}$. 
For convenience, we use $\varphi(\vec{\anivar})$ with $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ to denote an LIA formula $\varphi$ whose free variables are from $\{\anivar_1, \cdots, \anivar_k\}$. For an LIA formula $\varphi(\vec{\anivar})$ with $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ and $\myvec{t} = (t_1, \cdots, t_k)$ be a tuple of integer terms, we use $\varphi[\myvec{t}/\myvec{\anivar}]$ to denote the formula obtained by replacing (simultaneously) $\anivar_i$ with $t_i$ for every $i \in [k]$.
%For convenience, we usually assume that a set $X$ of variables is a vector. 
%, which is essentially Presburger

>>>>

\medskip
\noindent \emph{Strings and languages.}
%\paragraph*{Strings and languages.}
A string over $\Sigma$ is a (possibly empty) sequence of elements from $\Sigma$,
denoted by $u, v, w, \ldots$. An empty string is denoted by $\varepsilon$. We use $\Sigma_\varepsilon$ to denote $\Sigma \cup \{\varepsilon\}$. We write $\Sigma^*$ (resp., $\Sigma^+$) for the set of all (resp. nonempty) strings over $\Sigma$.
%Let 
For a string $u$, we use $|u|$ to denote the number of letters in $u$. In particular, $|\varepsilon|=0$. 
Moreover, for $a \in \Sigma$, let $|u|_a$ denote the number of occurrences of $a$ in $u$. 
Assume that $u=a_1\cdots a_{n}$ is nonempty and $i<j \in [n]$. %Then a \emph{position} of $u$ is a number $i \in [|u|]$ (Note that the first position is $1$, instead of  0). In addition, 
We let $u[i]$ denote $a_i$ and $u[i,j]$ for the substring %of $u$ starting from $i$ and ending with $j$ (i.e., 
$a_i\cdots a_j$. 
%\tl{check later for consistency}\zhilin{the indices start from 0, to be consistent with the semantics of substring and indexof}\zhilin{i do use it in section 4}
%
Let $u, v$ be two strings. We use $u \cdot v$ to denote the \emph{concatenation} of $u$ and $v$. A language $L$ over $\Sigma$ is a subset of strings.  
Let $L_1, L_2$ be two languages. Then the concatenation of $L_1$ and $L_2$, denoted by $L_1 \cdot L_2$,  is defined as $\{u \cdot v \mid u \in L_1, v \in L_2\}$. The union (resp. intersection) of $L_1$ and $L_2$, denoted by $L_1 \cup L_2$  (resp. $L_1 \cap L_2$), is defined as $\{u \mid u \in L_1 \mbox{ or } u \in L_2\}$ (resp. $\{u \mid u \in L_1 \mbox{ and } u \in L_2\}$). The complement of $L_1$, denoted by $\overline{L_1}$, is defined as $\{u \mid u \in \Sigma^*, u \not \in L_1\}$. 
The difference of $L_1$ and $L_2$, denoted by $L_1 \setminus L_2$, is defined as $L_1 \cap \overline{L_2}$. 
For a language $L$ and $n \in \Nat$, we define $L^n$ inductively as follows: $L^0= \{\varepsilon\}$, $L^{n+1} = L \cdot L^n$ for every $n \in \Nat$. 
Finally, define $L^* = \bigcup \limits_{n \in \Nat} L^n$.
%, that is, the string $w$ such that $|w|= |u| + |v|$ and for each $i \in [0, |u|-1]$, $w[i]= u[i]$, and for each $i \in [0,|v|-1]$, $w[|u|+i]=v[i]$.

@@

\medskip
\noindent 
\emph{Linear integer arithmetic and Parikh images.}
We will also use standard quantifier-free/existential \emph{linear integer arithmetic} (LIA) formulas, which are typically ranged over by $\phi, \varphi$, etc. 
For a set $\mathfrak{X}$ of variables, we use $\Phi(\mathfrak{X})$ to denote the set of quantifier-free LIA formulas whose free variables are from $\mathfrak{X}$. 
For convenience, we use $\varphi(\vec{\anivar})$ with $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ to denote an LIA formula $\varphi$ whose free variables are from $\{\anivar_1, \cdots, \anivar_k\}$. For an LIA formula $\varphi(\vec{\anivar})$ with $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ and $\myvec{t} = (t_1, \cdots, t_k)$ be a tuple of integer terms, we use $\varphi[\myvec{t}/\myvec{\anivar}]$ to denote the formula obtained by replacing (simultaneously) $\anivar_i$ with $t_i$ for every $i \in [k]$.
%For convenience, we usually assume that a set $X$ of variables is a vector. 
%, which is essentially Presburger

### 2023-05-12 2023-1-19/sections/recl.tex
section 5 and 6

%\subsubsection{Syntax}
In the sequel, we define string constraints with regex-counting and string-length function (abbreviated as RECL). The syntax of RECL is defined by the rules in Figure~\ref{fig:syntax}, where $x$ is a string variable, $\anivar$ is an integer variable, $a$ is a character from an alphabet $\Sigma$, and $m, n$ are integer constants. 
A RECL formula $\varphi$ is a conjunction of atomic formulas of the form $x \in \regex$, $x \not \in \regex$, or $t_1\ \op\ t_2$, where $\regex$ is a regular expression,  
$t_1$ and $t_2$ are integer terms, and $\op \in \{=, \neq, \le, \ge, <, >\}$. Atomic formulas of the form $x \in \regex$ or $x \not \in \regex$ are called \emph{regular membership} constraints and atomic formulas of the form $t_1\ \op\ t_2$ are called \emph{length} constraints. 
%
A regular expression $\regex$ is built from $\emptyset$, the empty string $\epsilon$, and characters $a$ by using concatenation  $\cdot$, union $+$, Kleene star $^*$, intersection $\cap$, complement $\bar{\mbox{ }}$, difference $\setminus$, counting $^{\{m,n\}}$ or $^{\{m,\infty\}}$. An integer term is built from constants $n$, variables $i$, and string lengths $|x|$, by operators $+$ and $-$.
% 
%is a quantifier-free formula that can be a regular membership $x\in\regex$, a quantifier-free Presburger formula $\alpha_1 \leq \alpha_2$, a negation of a formula, or a disjunction of two formulas. $\alpha$ is an integer term which can be an integer constant $m$, integer variable $i$, the length of the string term $|s|$, the minus of an integer term, and the plus of two integer terms. The regular expression $\regex$ is built on empty string $\epsilon$, constant letter $a\in \Sigma$. The supported regex operations involve concatenation $\cdot$, disjunction $+$, intersection $\times$, closure $*$, complement $C$, and bounded repetition $\{m, n\}$. We use $term(\varphi)$ to denote the set of terms and $strvar(\varphi)$ to denote the string variables occurring in $\varphi$. The \emph{literals} of $\varphi$ are $x\in \regex$ and $\alpha_1 \leq \alpha_2$. $x\in \regex$ is called \emph{regular literals} and $\alpha_1 \leq \alpha_2$ is called \emph{linear literals}.
\begin{figure}[h]
\vspace{-3mm}
  \centering
%  \begin{tabular}{l r}
$ \begin{array}{l l l r}
    \varphi & ::= & x\in \regex \mid t_1\  \op\ t_2 \mid  \varphi \wedge \varphi                                              & \mbox{formulas}            \\
    \regex &::= & \emptyset \mid \epsilon \mid a \mid \regex\cdot \regex \mid \regex+\regex \mid \regex^* \mid \regex \cap \regex \mid \overline{\regex} \mid \regex \setminus \regex \mid \regex^{\{m,n\}} \mid \regex^{\{m,\infty\}} \mid (e) & \mbox{regexes} \\
    t &::= & n \mid \anivar \mid  |x| \mid t - t \mid t + t                                                                    & \mbox{integer terms}
    \end{array}
  $
%  \end{tabular}
  \caption{Syntax of RECL }\label{fig:syntax}
\vspace{-3mm}
\end{figure}
Moreover, for $S \subseteq \Sigma$ with $S = \{a_1, \cdots, a_k\}$, we use $S$ as an abbreviation of $a_1 + \cdots + a_k$.

The definition of the semantics of RECL is standard. Nevertheless, for readibility, we include the definition of its semantics here.

@@

Let $\varphi$ be a RECL formula and $\svars(\varphi)$ (resp. $\ivars(\varphi)$) denote the set of string (resp. integer) variables occurring in $\varphi$. 
%
Then the semantics of formulas are defined with respect to a mapping $\theta: \svars(\varphi) \rightarrow \Sigma^* \uplus \ivars(\varphi) \rightarrow \Int$. 
The mapping $\theta$ can be naturally extended to the set of integer terms. For instance, $\theta(|x|) =|\theta(x)|$, $\theta(t_1 + t_2) = \theta(t_1) + \theta(t_2)$. 
A  mapping $\theta$ is said to satisfy $\varphi$, denoted by $\theta \models \varphi$, if one of the following holds: 
%\begin{itemize}
%\item 
$\varphi \equiv x \in \regex$ and $\theta(x) \in \Lang(\regex)$, 
%
%\item $\varphi \equiv x \not \in \regex$ and $\theta(x) \not \in \Lang(\regex)$, 
%
%\item 
$\varphi \equiv t_1\ \op\ t_2$ and $\theta(t_1)\ \op\ \theta(t_2)$, 
%
%\item 
$\varphi \equiv \varphi_1 \wedge \varphi_2$ and $\theta \models \varphi_1$ and $\theta \models \varphi_2$.
%\end{itemize}
A RECL formula $\varphi$ is said to be satisfiable if there is a mapping $\theta$ such that $\theta \models \varphi$. The satisfiability problem for RECL (abbreviated as $\reclsat$) is to decide whether a given RECL formula $\varphi$ is satisfiable.

>>>>

%\subsubsection{Syntax}
In the sequel, we define string constraints with regex-counting and string-length function (abbreviated as RECL). The syntax of RECL is defined by the rules in Figure~\ref{fig:syntax}, where $x$ is a string variable, $\anivar$ is an integer variable, $a$ is a character from an alphabet $\Sigma$, and $m, n$ are integer constants. 
A RECL formula $\varphi$ is a conjunction of atomic formulas of the form $x \in \regex$ or $t_1\ \op\ t_2$, where $\regex$ is a regular expression,  
$t_1$ and $t_2$ are integer terms, and $\op \in \{=, \neq, \le, \ge, <, >\}$. Atomic formulas of the form $x \in \regex$ are called \emph{regular membership} constraints and atomic formulas of the form $t_1\ \op\ t_2$ are called \emph{length} constraints. 
%
A regular expression $\regex$ is built from $\emptyset$, the empty string $\epsilon$, and characters $a$ by using concatenation  $\cdot$, union $+$, Kleene star $^*$, intersection $\cap$, complement $\bar{\mbox{ }}$, difference $\setminus$, counting $^{\{m,n\}}$ or $^{\{m,\infty\}}$. An integer term is built from constants $n$, variables $\anivar$, and string lengths $|x|$, by operators $+$ and $-$.
% 
%is a quantifier-free formula that can be a regular membership $x\in\regex$, a quantifier-free Presburger formula $\alpha_1 \leq \alpha_2$, a negation of a formula, or a disjunction of two formulas. $\alpha$ is an integer term which can be an integer constant $m$, integer variable $i$, the length of the string term $|s|$, the minus of an integer term, and the plus of two integer terms. The regular expression $\regex$ is built on empty string $\epsilon$, constant letter $a\in \Sigma$. The supported regex operations involve concatenation $\cdot$, disjunction $+$, intersection $\times$, closure $*$, complement $C$, and bounded repetition $\{m, n\}$. We use $term(\varphi)$ to denote the set of terms and $strvar(\varphi)$ to denote the string variables occurring in $\varphi$. The \emph{literals} of $\varphi$ are $x\in \regex$ and $\alpha_1 \leq \alpha_2$. $x\in \regex$ is called \emph{regular literals} and $\alpha_1 \leq \alpha_2$ is called \emph{linear literals}.
\begin{figure}[h]
\vspace{-3mm}
  \centering
%  \begin{tabular}{l r}
$ \begin{array}{l l l r}
    \varphi & ::= & x\in \regex \mid t_1\  \op\ t_2 \mid  \varphi \wedge \varphi                                              & \mbox{formulas}            \\
    \regex &::= & \emptyset \mid \epsilon \mid a \mid \regex\cdot \regex \mid \regex+\regex \mid \regex^* \mid \regex \cap \regex \mid \overline{\regex} \mid \regex \setminus \regex \mid \regex^{\{m,n\}} \mid \regex^{\{m,\infty\}} \mid (e) & \mbox{regexes} \\
    t &::= & n \mid \anivar \mid  |x| \mid t - t \mid t + t                                                                    & \mbox{integer terms}
    \end{array}
  $
%  \end{tabular}
  \caption{Syntax of RECL }\label{fig:syntax}
\vspace{-3mm}
\end{figure}
Moreover, for $S \subseteq \Sigma$ with $S = \{a_1, \cdots, a_k\}$, we use $S$ as an abbreviation of $a_1 + \cdots + a_k$.

%The definition of the semantics of RECL is standard. Nevertheless, for readibility, we include the definition of its semantics here.

@@

Let $\varphi$ be a RECL formula and $\svars(\varphi)$ (resp. $\ivars(\varphi)$) denote the set of string (resp. integer) variables occurring in $\varphi$. 
%
Then the semantics of  $\varphi$ is defined with respect to a mapping $\theta: \svars(\varphi) \rightarrow \Sigma^* \uplus \ivars(\varphi) \rightarrow \Int$ (where $\uplus$ denotes the disjoint union). 
Note that the mapping $\theta$ can be naturally extended to the set of integer terms. For instance, $\theta(|x|) =|\theta(x)|$, $\theta(t_1 + t_2) = \theta(t_1) + \theta(t_2)$. 
A  mapping $\theta$ is said to satisfy $\varphi$, denoted by $\theta \models \varphi$, if one of the following holds: 
%\begin{itemize}
%\item 
$\varphi \equiv x \in \regex$ and $\theta(x) \in \Lang(\regex)$, 
%
%\item $\varphi \equiv x \not \in \regex$ and $\theta(x) \not \in \Lang(\regex)$, 
%
%\item 
$\varphi \equiv t_1\ \op\ t_2$ and $\theta(t_1)\ \op\ \theta(t_2)$, 
%
%\item 
$\varphi \equiv \varphi_1 \wedge \varphi_2$ and $\theta \models \varphi_1$ and $\theta \models \varphi_2$.
%\end{itemize}
A RECL formula $\varphi$ is said to be satisfiable if there is a mapping $\theta$ such that $\theta \models \varphi$. The satisfiability problem for RECL (abbreviated as $\reclsat$) is to decide whether a given RECL formula $\varphi$ is satisfiable.
